### 2016-02-24 doc/draft.Rmd
added correlation coeff

```{r message=F,echo=FALSE,cache=T,results='hide',warning=FALSE}
xamp_roc<-"figure 1"
Xamp_roc<-"Figure 1"
miseq<-"figure 2"
Miseq<-"Figure 2"
hiseq<-"figure 3"
Hiseq<-"Figure 3"
qual<-"figure 4"
Qual<-"Figure 4"
freq<-"figure 5"
Freq<-"Figure 5"
lofreq<-"Figure 6"
Lofreq<-"figure 6"
lower<-"figure 7"
Lower<-"Figure 7"
dups<-"figure 8"
Dups<- "Figure 8"
Table<-"Table 1"
table<-"table 1"

@@

st_error="XXX"
false_p_phred<-round(false_p_phred,digits=1)
#csl: journal-of-virology.csl
```

@@

The ability to reliably distinguish true from false single nucleotide variants (SNV) is integral to accurate NGS based studies of viral diversity. The accuracy of any SNV calling pipeline can be described in terms of sensitivity and specificity. Sensitivity is the proportion of true variants properly identified; while, specificity is the proportion of true negatives properly identified. Specificity is often reported as 1-specificity or the false positive rate. The false positive rate gives the ratio of false positive variants to true negatives. Both a sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and only true variants are found.

There is an obvious trade off between sensitivity and specificity. Improved sensitivity often requires less stringent criteria in variant calling which reduces specificity. Conversely, increased stringency can improve specificity but often reduces sensitivity. This relationship can be represented as a receiver operating characteristic (ROC) curve (`r Xamp_roc`). An ROC curve plots the sensitivity of an assay along the y axis and 1-specificity along the x axis. To make an ROC curve for a variant calling pipeline one must test the assay on known data. The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score. In a controlled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut-offs (`r Xamp_roc` A). These points are then used to construct the curve (`r Xamp_roc` B). A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.

@@

A comprehensive comparison of SNV calling approaches is beyond the scope of this work. Here, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving accuracy. In the process we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.

DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome. Because it is clonal, the sequence of the control is known with a high degree of confidence and so any nonconsensus base is indicative of an error. Additionally, the control and experimental samples are processed together and are assumed to have identical noise characteristics. DeepSNV then applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies significantly above the noise found in the plasmid control. Like many variant calling methods, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification. We created defined mixtures of two influenza strains, WSN33 and PR8, to determine the accuracy of DeepSNV when applied to RT-PCR amplified RNA virus populations.

WSN33 and PR8 were plaque purified and sanger sequenced. cDNA genomes from both viruses were mixed serially such that WSN33 was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A). These viruses differ at 491 positions (primer sites were excluded from analysis), providing 491 true positives in each dilution. On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0. Under our experimental conditions and with a slightly more stringent p value  ( p=0.01 vs. p=0.05) we found a moderate reduction in sensitivity  (0.830 for variants at 0.63\% and 0.14 for variants at 0.16\%). The specificity was above 0.998 in all dilutions. While this drop in specificity appears minor, it is important to keep in mind that even a false positive rate of 0.002 results in 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb`kb of the influenza genome we analyzed (3 variants per genomic position).

@@

Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions. The virus populations found in patient-derived samples are  less divergent than WSN33 and PR8 (cite). Also the nucleic acid concentration in patient samples is much lower than that found in samples grown in cell culture (data not shown). To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background. We sequenced these mutants on the Illumina Miseq platform and confirmed that no additional mutations >1\% frequency had arisen between transfection and the passage 1 stock. Accross all 20 clones 4 additional mutations were found (frequencies 1.2%-3.7%). We also quantified the genome copy number of each stock using quantitative PCR. We then mixed equal genome equivalents of these 20 viruses, to generate an artificial population with $10^5$ copies per microliter with each mutation  present at 5\% frequency. This population was serially diluted into a stock of WSN33, generating artificial populations with all 20 mutations present a 2, 1, 0.5, 0.2, and 0.1\% frequency (`r Hiseq` A). We then serially diluted these populations into basal media to obtain mixtures with lower nucleic acid input. The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown). We sequenced the samples on the Illumina Hiseq platform, and variants were called using DeepSNV. We also processed and sequenced the WSN33 stock in duplicate to control for any background mutations.

@@

The samples with the greatest input concentration and variant frequencies of 1\% and greater had the highest sensitivity (`r hiseq` B and C). Sensitivity dropped at the 0.5\% frequency, and in general was lower than that seen previously in the cell culture samples. Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was lower by 1,000 fold.  The reduced sensitivity is likely due to the fact that we expect only 7,000 copies of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above. During library prep, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.

The specificity was greater than 0.99 in all the samples with $10^5$ genomes per microliter. As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome (again primer sites were excluded). Increased specificity can be achieved by applying a more stringent p-value cutoff. However, as shown by the ROC curves in `r hiseq`, this move towards the y axis would markedly reduce sensitivity. Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.

@@

In addition to distinguishing true from false SNV, DeepSNV was able to estimate the frequency of the true positive variants with reasonable accuracy (`r Freq`). However, while the median frequencies are near the expected values, there is a fair spread in each sample (mean standard error of `r st_error`). This error should be kept in mind when employing down stream analysis that depend on frequency measurements such as diversity metrics like Shannon's Entropy and Simpson's D, as well as inter-sample comparison of population structure .

>>>>

```{r message=F,echo=FALSE,cache=T,results='hide',warning=FALSE}
xamp_roc<-"figure 1"
Xamp_roc<-"Figure 1"
miseq<-"figure 2"
Miseq<-"Figure 2"
hiseq<-"figure 3"
Hiseq<-"Figure 3"
qual<-"figure 4"
Qual<-"Figure 4"
freq<-"figure 5"
Freq<-"Figure 5"
lofreq<-"Figure 6"
Lofreq<-"figure 6"
lower<-"figure 7"
Lower<-"Figure 7"
dups<-"figure 8"
Dups<- "Figure 8"
Table<-"Table 1"

@@

st_error="XXX"
false_p_phred<-round(false_p_phred,digits=1)
corr_coef<-round(corr_coef,digits=4)
mean_fold<-round(mean_fold,digits=2)*100
#print(interest)
#csl: journal-of-virology.csl
```

```{r,echo=F}
#Is it clear that I'm only talking about the 10^5 samples until we get to the lower input section
```

@@

The ability to reliably distinguish true from false single nucleotide variants (SNV) is integral to accurate NGS based studies of viral diversity. The accuracy of any SNV calling pipeline can be described in terms of its sensitivity and specificity. Sensitivity is the proportion of true variants that are properly identified; while, specificity is the proportion of true negatives that properly identified. Specificity is often reported as 1-specificity or the false positive rate. The false positive rate gives the ratio of false positive variants to true negatives. A sensitivity and specificity of 1 describe perfect accuracy in which all the true variants are found and only true variants are found.

There is an obvious trade off between sensitivity and specificity. Improved sensitivity often requires less stringent criteria in variant calling which reduces specificity. Conversely, increased stringency can improve specificity but often reduces sensitivity. This relationship can be represented as a receiver operating characteristic (ROC) curve (`r Xamp_roc`). An ROC curve plots the sensitivity of an assay along the y axis and 1-specificity along the x axis. A variant calling pipeline must be tested against known data in order to construct an ROC curve. The outcomes can then be stratified according to a metric that quantifies the probability that a given variant is real; often this is a p value or quality score. In a controlled benchmakring experiment, all true variants are known, and the sensitivity and specificity can be calculated at different cut-offs (`r Xamp_roc` A). These points are then used to construct the curve (`r Xamp_roc` B). A perfect ROC curve in which all the true positives can be separated from all false positives is a right angle that follows the upper left perimeter of the plot.

@@

A comprehensive comparison of SNV calling approaches is beyond the scope of this work. Instead, we robustly benchmark one variant caller, DeepSNV, and highlight approaches for improving accuracy. In the process we demonstrate the importance of robustly benchmarking any variant calling method under the experimental conditions to which it is being applied.

DeepSNV is a variant calling algorithm that uses a clonal, plasmid-derived, control to estimate local error rates across the genome. Because it is clonal, the sequence of the control is known with a high degree of confidence and so any nonconsensus base is indicative of an error. Additionally, the control and experimental samples are processed together and are assumed to have identical noise characteristics. DeepSNV then applies a hierarchical binomial model at each genomic position and identifies true variants as those with frequencies significantly above the noise found in the plasmid control. Like many variant calling methods, the accuracy of DeepSNV was initially determined using samples that required minimal PCR amplification. However, the accuracy had not been tested when applied to whole genome sequencing of RT-PCR amplified viral population. We created defined mixtures of two influenza strains, WSN33 and PR8, to determine the accuracy of DeepSNV under these conditions.

WSN33 and PR8 samples were plaque purified and sanger sequenced. cDNA genomes from both viruses were mixed serially such that WSN33 was present at frequencies of 5, 2.5, 1.25, 0.63 and 0.16\% (`r Miseq` A). These viruses differ at 491 positions (primer sites were excluded from analysis), providing 491 true positives in each dilution. On plasmids subjected to limited PCR, DeepSNV identified known variants at 0.1\% frequency with a sensitivity of 0.860 and  specificity of 1.0. Under our experimental conditions and with a slightly more stringent p value  ( p=0.01 vs. p=0.05) we found a moderate reduction in sensitivity  (0.830 for variants at 0.63\% and 0.14 for variants at 0.16\%). The specificity was above 0.998 in all dilutions. While this drop in specificity appears minor, it is important to keep in mind that even a false positive rate of 0.002 results in 78 false positives when applied across the `r miseq_vars` potential variants present in the `r miseq_kb`kb of the influenza genome we analyzed (3 variants per genomic position).

@@

Although our first benchmarking experiment validated our ability to accurately detect rare variants in influenza populations, the experiment was run under relatively artificial conditions. The virus populations found in patient-derived samples are  less divergent than WSN33 and PR8 (cite). Also the nucleic acid concentration in patient samples is much lower than that found in samples grown in cell culture (data not shown). To mimic these conditions, we generated 20 viral clones, each with a single point mutation in the WSN33 background. We sequenced these mutants on the Illumina Miseq platform to account for any additional mutations that might have arisen between transfection and the passage 1 stock. Four additional mutations were found above 1% frequency and were masked from analysis (frequencies 1.2%-3.7%). We also quantified the genome copy number of each stock using quantitative PCR. We then mixed equal genome equivalents of these 20 viruses, to generate an artificial population with $10^5$ copies per microliter with each mutation  present at 5\% frequency. This population was serially diluted into a stock of WSN33, generating artificial populations with all 20 mutations present a 2, 1, 0.5, 0.2, and 0.1\% frequency (`r Hiseq` A). We then serially diluted these populations into basal media to obtain mixtures with lower nucleic acid input. The range ($10^3-10^5$ copies per microliter) matches the inputs typically found in patient-derived samples (data not shown). We sequenced the samples on the Illumina Hiseq platform, and variants were called using DeepSNV. We also processed and sequenced the WSN33 stock in duplicate to control for any background mutations.

@@

The samples with the greatest input concentration and variant frequencies of 1\% and greater sensitivities of 0.850 or greater (`r Hiseq` B and C). Unlike what was seen in cell culture samples abover, sensitivity dropped drastically at 0.5\% frequency. Typically our samples grown in cell culture have  at least $10^8$ genomes per microliter; whereas, the nucleic acid concentration in these samples was lower by 1,000 fold.  The reduced sensitivity is likely due to the fact that we expect only 7,000 copies of each variant genome in the 0.5\% frequency samples compared to the equivalent of 62,500,000 variant genomes in the 0.63\% sample from the cell culture experiment above. During library preperation, a disproportionate amount of the 2.0\% sample was lost compared to the others, which likely accounts for the reduced sensitivity relative to the 1.0\% sample.

The specificity was greater than 0.99 in all the samples with $10^5$ genomes per microliter. As above, while 0.99 specificity appears robust, a false positive rate of 0.01 results in over 200 false positive variants when applied to the `r hiseq_vars` potential variants in the influenza genome (again primer sites were excluded). Increased specificity can be achieved by applying a more stringent p-value cutoff. However, as shown by the ROC curves in `r Hiseq`, this move towards the y axis would markedly reduce sensitivity. Our data demonstrate that with moderate concentrations of input nucleic acid, even statistically significant p-values from a robust variant caller are not sufficient to accurately separate true from false positive variants.

@@

In addition to distinguishing true from false SNV, DeepSNV was able to estimate the frequency of the true positive variants (`r Freq`). Although the median of the measured frequencies match the expetect values we found substantial spread about the median and the overall the fit was quite poor ( R^2^=`r corr_coef`). On average the percent difference between the measured frequency an the expected was `r mean_fold`\%. This error should be kept in mind when employing down stream analysis that depend on frequency measurements such as diversity metrics like Shannon's Entropy and Simpson's D, as well as inter-sample comparison of population structure .
